Online supplement accompanying 'The Locally Gaussian Partial Correlation

### 2020-09-02 lgpc_JBES_supp.Rmd
Implementing comments from Dag and one referee

All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally Gaussian joint density

\begin{equation}
\psi(\z) = \frac{1}{(2\pi)^{(d_1+d_2+d_3)/2}|\fSigma(\z)|^{1/2}} \exp \left\{-\frac{1}{2}(\z - \fmu(\z))^T\fSigma^{-1}(\z)(\z - \fmu(\z))\right\}
(\#eq:localgaussian0)
\end{equation}
and the locally Gaussian *conditional* density

@@

Let $\D = \textrm{diag}\left\{\fSigma_{12|3}\right\}$. The local Gaussian partial correlation matrix can be expressed as the lower left hand block matrix of $\D^{-1/2}\fSigma_{12|3}\D^{-1/2}$, and so the LGPC matrix is thus given as the upper right hand block of the matrix
\begin{equation}
\D^{-1/2}\fSigma_{12|3}\D^{-1/2} = \D^{-1/2}\begin{pmatrix} \fSigma_{11} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{12} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{32} \\ \fSigma_{21} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{22} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{32}\end{pmatrix}\D^{-1/2}.
(\#eq:matrixdef)
\end{equation}
Let us consider this expression for a few special cases.

@@

Testing for conditional independence between $\Z_1$ and $\Z_2$ given $\Z_3$ is in our context the same as testing for zero local partial correlation between all pairs of variables, a test for which is presented in the main article and discussed in technical detail later in this supplement. One can also, based on \eqref{eq:offdiag1} and \eqref{eq:offdiag2}, test for conditional independence between combinations of components given $\Z_3$. Equations \eqref{eq:offdiag1}--\eqref{eq:offdiag3} above give explicit expressions for the conditional correlation matrix of $\Z_1$ and $\Z_2$ given $\Z_3$.

\item[\textbf{c)}] \textbf{Let} $\bm{d_1 = d_2 = 1}$ \textbf{and let} $\bm{d_3}$ \textbf{be arbitrary}. This is the case usually treated in the literature, also the focus of the discussions in the main article. In this case
$$\fSigma^{11} = \begin{pmatrix} 1 & \rho_{12} \\ \rho_{21} & 1 \end{pmatrix}, \fSigma_{13} = \left\{\rho_{1i}^{1,3}\right\}, i=1,\ldots,d_3; \fSigma_{23} = \left\{\rho_{2i}^{2,3}\right\}; i=1,\ldots,d_3,$$
hence $\begin{pmatrix} \fSigma_{13} \fSigma_{23}\end{pmatrix}^T$ is of dimension $2\times d_3$, whereas $\begin{pmatrix} \fSigma_{31} \fSigma_{32}\end{pmatrix}$ is of dimension $d_3\times 2$ such that
$$\begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix} \fSigma_{33}^{-1} \begin{pmatrix} \fSigma_{31} & \fSigma_32 \end{pmatrix}$$
is of dimension $2\times2$. However, as is immediately seen, no formulas like \eqref{eq:offdiag1}--\eqref{eq:offdiag3} will appear; they will be much more complicated and include sums and products of various correlations.

@@

\item[\textbf{(ii)}] If $\bm{d_1 = d_2 = 1}$ \textbf{and} $\bm{d_3 > 1}$, we estimate all correlations in a pairwise manner as laid out in the main article. The limiting distribution of the (scalar) local Gaussian partial correlation is in general given by 
\begin{equation}
\sqrt{nb_n^2}\left(\widehat{\alpha}(z))-\alpha(z)\right)  \stackrel {{\cal L}}{\to} {\cal N}(0,\nabla g(\frho)^{T} {\fSigma} \nabla g (\frho)),
\end{equation}
see eq. (22) in the main article, and see Theorem 2 in Otneim and Tjøstheim (2018) for a definition of $\fSigma$. The computation of $\nabla g$ is outlined in Section \ref{app-asvar}.
\end{itemize}

@@

\begin{equation}
\int K_b(v-z_i, w-z_j)\big[\psi(v,w; \rho_b(z_i, z_j)) - \log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w)\big] \di v \di w,
(\#eq:penalty)
\end{equation}
where $f$ is the density function of $(Z_i, Z_j)$. Note in particular that $\rho_{0,b} \rightarrow \rho_0$ as $b\rightarrow0$, which is the true local correlation parameter. The reader may consult @tjostheim2013local for more details and regularity conditions in this regard. The corresponding estimate $\rho_{n, b}(z_i,z_j)$ is obtained by maximizing the log-likelihood in \@ref(eq:loglikelihood). Note that, writing $\rho$ for $\rho_b$,
\begin{equation}
L_{n,b}(\rho) = \frac{1}{n}\sum_{t=1}^n L^{(b)}(Z_{i,t}, Z_{j,t}; \rho),
\end{equation}
where
\begin{align}
L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad \int K_b(v-z_i, w- z_j)\rho_b(v,w;\psi_b(z_i, z_j)) \, \di v \di w.
\label{eq:A16}
\end{align}

@@

\begin{itemize}
\item[\textbf{A1}] $(\Omega, \mathcal{F}, \mathbb{P})$ is a complete probability space on which is defined a sequence of random functions $\{L_{n,b}: \Omega\rightarrow\overline{\mathbb{R}}\}$, where $\overline{\mathbb{R}}$ is the extended real line. The parameter set $\Theta$ is compact (trivially fulfilled in our case because $\Theta = [-1,1]$). The random process $\{Z_{i,t}, Z_{j,t}\}$ is $\mathcal{F}_t$-measurable, stationary and $\alpha$-mixing with mixing coefficients $\alpha_k = O(\alpha^k)$, $\alpha \in (0,1)$ for all $t$, and with a bivariate density having support in all of $\mathbb{R}^2$. Here $\mathcal{F}_t$ is the $\sigma$-algebra generated by $\{Z_{i,s}, Z_{j,s}, s \leq t\}$.
\item[\textbf{A2}] The parameter value $\rho_{0,b}$ is the unique maximizer of $\E(L_{n,b}(\rho)): \Theta \rightarrow \overline{\mathbb{R}}$, with $\E(L_{n,b}(\rho))$ is defined in \eqref{eq:penalty}.
\end{itemize}

@@

The proof of Theorem \@ref(thm:A1) is mostly identical to the proof of Theorem 2 of @lacal2018estimating with trivial modifications to take into account the scalar nature of the parameter $\rho_{ij}$ in the present case. A distributional limit theorem analogous to Theorem 2.2 in @lacal2018estimating can also be rephrased and proved as in that paper, but we refrain from doing this given the distributional results already stated in Section 3 of the main part of the paper.

@@

The second term in $\Jb$ can similarly be written as $II_J = B(\psi(\z, \theta_h(\z)) - f(\z)) + o(b^2)$, where $B = \nabla \fv(\z, \theta_b(\z))$ are second derivatives of the local likelihood function with respect to the local correlations, and $\psi(\z, \theta_b(\z)) - f(\z) = O(b^2)$ according to @hjort1996locally. Hence, it follows that the leading term of $\Jb^{-1}\Mb(\Jb^{-1})^T$ is $O(b^{-2})$, which again means that the convergence rate in eq. \@ref(eq:fixed-h) above must be modified accordingly in order to balance the convergence in eq. (20) in the main article. Extracting an analytic expression for the leading term of the asymptotic covariance matrix is, as mentioned by @tjostheim2013local,  possible by means of symbolic manipulation software, but the expression is very complicated and of little practical use. See also the corresponding proof of Theorem 3 in @tjostheim2013local.

@@

Proving joint asymptotic normality of the local Gaussian correlations with marginally normally distributed variables $(Z_1, \ldots, Z_p) = \left(\Phi^{-1}(F_1(X_1)), \ldots, \Phi^{-1}(F_p(X_p))\right)$, where $F_j$ is the cdf of $X_j$ for $i = 1,\ldots,p$, relies on proving asymptotic normality for the variables
$$Y_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\Z_t - \z)\fv(\Z_t, \frho_0),$$
where $\fv(\cdot) = \partial L(\cdot)/\partial\frho$ in the same way as in Section \@ref(app-loccor) in this supplement. This has been done by @otneim2017locally in the iid case using arguments from @schervish1995theory, and by @otneim2017conditional in the $\alpha$-mixing case, using arguments from @fan2008nonlinear. We need here to prove asymptotic normality of the variables

$$\hY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\hZ_t - \z)\fv(\hZ_t,\frho_0),$$
with $\hnZ_t = \Phi^{-1}(F_n(\X_t))$, and $F_n(\X_t)$ is the application of the marginal empirical distribution functions to the corresponding components of the vector $\X_t$.

In order to ease notation we state the proof in the simplified case only, where the local correlations are estimated using *pairs* $(i,j)$ of variables. It can be extended to the trivariate case by adding the appropriate terms in the derivations below, without changing the final conclusion. Write $\hY_n(\z) = Y_n(\z) - (Y_n(\z) - \hY_n(\z))$ and do a Taylor expansion of $Y_n(\z)$ around $F_n(\X_t)$. Since we use product kernels, we then need derivatives of (assuming without loss of generality that $K(x) = K(-x)$)

@@

In the following we omit the index $i=1,2$. We need a bound on the supremum of $|F_n(X_t) - F(X_t)|$ and $|F_n(x) - F(x)|$ as $n \rightarrow \infty$. According to @arcones1995law, the best law of the iterated logarithm has been obtained by @rio1995functional. One statement of this is as follows:

Let $f(x) \in L_p$, $2<p\leq\infty$, $\sum\alpha_k k^{2/(p-2)} < \infty$, then the sequence $f(X_t)$ satisfies the compact LIL (*law of the iterated logarithms*), i.e.
$$\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{t=1}^n(f(X_t) - E[f(X_t)]) \right\}_{n=1}^{\infty}$$
is relatively compact and its limit set is $[-\sigma, \sigma]$ where
$$\sigma^2 = \textrm{Var}\Big(f(X_t)\Big) + 2\sum_{s=1}^{\infty}\textrm{Cov}\Big(f(X_t), f(X_s)\Big).$$
For a fixed $x$, take $f(X_t) = I(X_t\leq x)$. Then E$(f(X_t)) = F(x)$ and $f(x)\in L_p$ for all $p$, and
$$\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{s=1}^n \left(I(X_s \leq x) - F(x) \right) \right\}$$
has its limit set in $[-\sigma, \sigma]$. Here 
$$\textrm{Var}(I(X_t \leq x)) = F(x)(1-F(x)) \leq \frac{1}{4},$$
and according to Davydov's Lemma (see e.g. @fan2008nonlinear p. 278),
$$\big| \textrm{Cov}(I(X_t \leq x), I(X_s \leq x))\big| \leq 8 \|I(X_t \leq x)\|_p \|I(X_s \leq x)\|_q \alpha(|s-t)^{1-p^{-1}-q^{-1}}, \,\,\, p,q>1, \,\,\, \frac{1}{p}+\frac{1}{q} < 1,$$
with $\| I(X_t \leq x)\|_p = \left(\textrm{E}|I(X_t \leq x)|^p\right)^{1/p}$. We may for example take $p=q=4$, leading to $\| I(X_t \leq x) \|_4 = F(x)^{1/4} \leq 1$.

Since we have assumed exponential mixing (much weaker conditions suffice), then Rio's condition $\sum \alpha_k k^{2/(p-2)} < \infty$ is satisfied, and the bounds of $-\sigma$ and $\sigma$ can be made independent of $x$.

Having 
$$F_n(x) = \frac{1}{n}\sum_{t=1}^n I(X_t \leq x)$$
means that
$$n^{\frac{1}{2}}[2\log \log n]^{-\frac{1}{2}} (F_n(x) - F(x))$$
has its limits in $[-\sigma, \sigma]$ or that the supremum of $|F_n(x) - F(x)|$ is almost surely bounded by $n^{-1/2}(2\log\log n)^{1/2}$.

This immediately implies that
$$\frac{1}{n}\sum_{t=1}^n\left(F_n(X_t) - F(X_t)\right)$$
is almost surely (and consequentially in probability) bounded by $n^{-\frac{1}{2}}(2 \log \log n)^{\frac{1}{2}}$ which for all conceivable bandwidths  is faster than the convergence rates of the CLT for $Y_n(\z)$, which is $(nb^2)^{-\frac{1}{2}}$.

@@

We can use the bootstrap theory by @lacal2018estimating more or less directly, which in turn is based on Gon\c{c}alves and White [-@gonccalves2002bootstrap; -@gonccalves2004maximum], but we spell out some details  below. First, we need to specify the analogue of the conditions A1--A5 by @lacal2018estimating, where we adapt their notation to our situation by replacing the parameter $\theta$ with $\rho_{ij}$, and the time series $\{X_t, Y_t\}$ by our multivariate observations $\{\Z_t\}$, possibly a multivariate time series. Since $\rho = \rho_{ij}$ is estimated using the two corresponding data vectors only, we can, for the time being and without loss of generality, assume that the observation series is bivariate: ${\Z_t} = \{Z_{i,t}, Z_{j,t}\}$. Note further that it suffices to consider the true pseudo-observations $\{\Z_t\}$ instead of the empirical pseudo observations $\{\hZ_t\}$ in the following because the difference between them is asymptotically negligible according to Theorem 3.2 in the main article.

To state the bootstrap results we need a few more concepts and three additional conditions. We block bootstrap the entire vector series $\{\Z_t\}$. This automatically leads to the same block-bootstrap for any pair of components which we have denoted by $\{Z_{i,t}, Z_{j,t}\}$ above. @lacal2018estimating use both the block bootstrap as well as the stationary bootstrap, and the results below hold irrespective of this particular choice. We will therefore not distinguish between them here, but rather refer the reader to @lacal2018estimating for further details on this distinction. Bootstrapped quantities will be denoted with a star ($*$) in the following.

@@

\begin{cor}
Assume the conditions of Theorem \ref{thm:E1}. Let $\alpha_{n,b}(i,j|\Z_3) \stackrel{\textrm{def}}{=} \widehat \rho(Z_{1i}, Z_{2j}|\Z_3)$ be the $i,j$th element of the estimated partial correlation matrix of $(\Z_1, \Z_2)$ given $\Z_3$. This depends on scalar coordinates $(z_1, z_2)$ and a vector coordinate $\z_3$. Note that $\alpha_{n,b}(i,j|\Z_3)$ is a continuous function of a vector of local correlations according to one of the alternatives (i)', (ii)--(v) in Section \ref{chap:estimation}. Then $\alpha_{n,b}(i,j|\Z_3) - \alpha^*_{n,b}(i,j|\Z_3) \buildrel {\mathbb{P}_{\omega}^{*},\mathbb{P}} \over {\longrightarrow} 0$. 
\label{cor:E3}
\end{cor}

@@

It should be carefully noted that it is not entirely obvious what the order of the term $C_n(A_b)$ is as $\bb = (b_i,b_j) = (b,b) \to \bm{0}$ and $n \to \infty$. Looking at the expressions \@ref(eq:t3) and \@ref(eq:t4), it is the term \@ref(eq:t3) that causes potential trouble due to the presence of $b_i^{-1}$ and $b_j^{-1}$ in the two kernel function terms in the product kernel. These are cancelled out by the integration over $(z_i,z_j)$, however. This means that the true convergence rate for $T_{n,b}-T$ is $n^{-1/2}$. This may be somewhat surprising since the convergence rate of $\frho_{n,b}(\z_{ij})-\frho(\z_{ij})$ is $(nb_1b_2)^{-1/2}$, but this is a well-known phenomenon for functionals of type $T_{n,b}$. In @lacal2018estimating the convergence rate is changed from $(nb_1^3b_2^3)^{-1/2}$ to $(nb_1^2b_2^2)^{-1/2}$. These results are consistent with similar results for test functionals for independence in @joe1989estimation.

@@

Once a bootstrap realization $\{\Z_{1,t}^*,\Z_{2,t}^*,\Z_{3,t}^*\}$ is created, following @su2008nonparametric, these realizations are now taken as the original observations in the local likelihood estimation procedure. For a fixed bandwidth $b$ the population values, following the local Gaussian approach, can be taken as the solution of the integral equation (2) in @tjostheim2013local; see also corresponding equations in e.g. Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional]. Note that in the present paper we have largely operated with two alternatives, the general vector case with inference via the pairwise local likelihood, or the the purely scalar case with three coordinates for the local Gaussian correlation. In both cases we have normalized the processes so that they have standard normal marginals.

>>>>

All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally Gaussian joint density $\psi(\z,\fv)$ at the point $\fv = \z$,

\begin{equation}
\psi(\z) = \frac{1}{(2\pi)^{(d_1+d_2+d_3)/2}|\fSigma(\z)|^{1/2}} \exp \left\{-\frac{1}{2}(\z - \fmu(\z))^T\fSigma^{-1}(\z)(\z - \fmu(\z))\right\}
(\#eq:localgaussian0)
\end{equation}
and the locally Gaussian *conditional* quantity

@@

Let $\D = \textrm{diag}\left\{\fSigma_{12|3}\right\}$. The local Gaussian partial correlation matrix can be expressed as the lower left hand block matrix of $\D^{-1/2}\fSigma_{12|3}\D^{-1/2}$, and so the LGPC matrix is thus given as the lower left hand block of the matrix
\begin{equation}
\D^{-1/2}\fSigma_{12|3}\D^{-1/2} = \D^{-1/2}\begin{pmatrix} \fSigma_{11} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{12} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{32} \\ \fSigma_{21} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{22} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{32}\end{pmatrix}\D^{-1/2}.
(\#eq:matrixdef)
\end{equation}
Let us consider this expression for a few special cases.

@@

Testing for conditional independence between $\Z_1$ and $\Z_2$ given $\Z_3$ is in our context the same as testing for zero local partial correlation between all pairs of variables, a test for which is presented in the main article and discussed in technical detail later in this supplement. One can also, based on \eqref{eq:offdiag1} and \eqref{eq:offdiag2}, test for conditional independence between combinations of components given $\Z_3$. Equations \eqref{eq:offdiag1}--\eqref{eq:offdiag3} above give explicit expressions for the conditional correlation matrix of $\Z_1$ and $\Z_2$ given $\Z_3$, where estimated values give a measure of local conditional dependence between the components.

\item[\textbf{c)}] \textbf{Let} $\bm{d_1 = d_2 = 1}$ \textbf{and let} $\bm{d_3}$ \textbf{be arbitrary}. This is the case usually treated in the literature, also the focus of the discussions in the main article. In this case
$$\fSigma^{11} = \begin{pmatrix} 1 & \rho_{12} \\ \rho_{21} & 1 \end{pmatrix}, \fSigma_{13} = \left\{\rho_{1i}^{1,3}\right\}, i=1,\ldots,d_3; \fSigma_{23} = \left\{\rho_{2i}^{2,3}\right\}; i=1,\ldots,d_3,$$
hence $\begin{pmatrix} \fSigma_{13} & \fSigma_{23}\end{pmatrix}^T$ is of dimension $2\times d_3$, whereas $\begin{pmatrix} \fSigma_{31} & \fSigma_{32}\end{pmatrix}$ is of dimension $d_3\times 2$ such that
$$\begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix} \fSigma_{33}^{-1} \begin{pmatrix} \fSigma_{31} & \fSigma_{32} \end{pmatrix}$$
is of dimension $2\times2$. However, as is immediately seen, no formulas like \eqref{eq:offdiag1}--\eqref{eq:offdiag3} will appear; they will be much more complicated and include sums and products of various correlations.

@@

\item[\textbf{(ii)}] If $\bm{d_1 = d_2 = 1}$ \textbf{and} $\bm{d_3 > 1}$, we estimate all correlations in a pairwise manner as laid out in the main article. The limiting distribution of the (scalar) local Gaussian partial correlation is in general given by 
\begin{equation}
\sqrt{nb_n^2}\left(\widehat{\alpha}(z))-\alpha(z)\right)  \stackrel {{\cal L}}{\to} {\cal N}(0,\nabla g(\frho)^{T} {\fLambda} \nabla g (\frho)),
\end{equation}
see eq. (22) in the main article, and see Theorem 2 in Otneim and Tjøstheim (2018) with $\fSigma = \fLambda$ for a definition of $\fLambda$. The computation of $\nabla g$ is outlined in Section \ref{app-asvar}.
\end{itemize}

@@

\begin{equation}
\int K_b(v-z_i, w-z_j)\big[\psi(v,w; \rho_b(z_i, z_j)) - \log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w)\big] \di v \di w,
(\#eq:penalty)
\end{equation}
where $f$ is the density function of $(Z_i, Z_j)$. Note in particular that $\rho_{0,b} \rightarrow \rho_0$ as $b\rightarrow0$, which is the true local correlation parameter. The reader may consult @tjostheim2013local for more details and regularity conditions in this regard. The corresponding estimate $\widehat \rho = \rho_{n, b}(z_i,z_j)$ is obtained by maximizing the log-likelihood in \@ref(eq:loglikelihood). Note that, writing $\rho$ for $\rho_b$,
\begin{equation}
L_{n,b}(\rho) = \frac{1}{n}\sum_{t=1}^n L^{(b)}(Z_{i,t}, Z_{j,t}; \rho),
\end{equation}
where
\begin{align}
L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\rho_b(v,w;\psi_b(z_i, z_j)) \, \di v \di w.
\label{eq:A16}
\end{align}

@@

\begin{itemize}
\item[\textbf{A1}] $(\Omega, \mathcal{F}, \mathbb{P})$ is a complete probability space on which is defined a sequence of random functions $\{L_{n,b}: \Omega\rightarrow\overline{\mathbb{R}}\}$, where $\overline{\mathbb{R}}$ is the extended real line. The parameter set $\Theta$ is compact (trivially fulfilled in our case because $\Theta = [-1,1]$). The random process $\{Z_{i,t}, Z_{j,t}\}$ is $\mathcal{F}_t$-measurable, stationary and $\alpha$-mixing with mixing coefficients $\alpha_k = O(\alpha^k)$, $\alpha \in (0,1)$ for all $t$, and with a bivariate density having support in all of $\mathbb{R}^2$. Here $\mathcal{F}_t$ is the $\sigma$-algebra generated by $\{Z_{i,s}, Z_{j,s}, s \leq t\}$.
\item[\textbf{A2}] The parameter value $\rho_{0,b}$ is the unique maximizer of $\E(L_{n,b}(\rho)): \Theta \rightarrow \overline{\mathbb{R}}$, with $\E(L_{n,b}(\rho))$ defined in \eqref{eq:penalty}.
\end{itemize}

@@

The proof of Theorem \@ref(thm:A1) is mostly identical to the proof of Theorem 2.1 of @lacal2018estimating with trivial modifications to take into account the scalar nature of the parameter $\rho_{ij}$ in the present case. A distributional limit theorem analogous to Theorem 2.2 in @lacal2018estimating can also be rephrased and proved as in that paper, but we refrain from doing this given the distributional results already stated in Section 3 of the main part of the paper.

@@

The second term in $\Jb$ can similarly be written as $II_J = B(\psi(\z, \frho_b(\z)) - f(\z)) + o(b^2)$, where $B = \nabla \fv(\z, \frho_b(\z))$ are second derivatives of the local likelihood function with respect to the local correlations, and $\psi(\z, \frho_b(\z)) - f(\z) = O(b^2)$ according to @hjort1996locally. Hence, it follows that the leading term of $\Jb^{-1}\Mb(\Jb^{-1})^T$ is $O(b^{-2})$, which again means that the convergence rate in eq. \@ref(eq:fixed-h) above must be modified accordingly in order to balance the convergence in eq. (20) in the main article. Extracting an analytic expression for the leading term of the asymptotic covariance matrix is, as mentioned by @tjostheim2013local,  possible by means of symbolic manipulation software, but the expression is very complicated and of little practical use. See also the corresponding proof of Theorem 3 in @tjostheim2013local.

@@

Proving joint asymptotic normality of the local Gaussian correlations with marginally normally distributed variables $(Z_1, \ldots, Z_p) = \left(\Phi^{-1}(F_1(X_1)), \ldots, \Phi^{-1}(F_p(X_p))\right)$, where $F_j$ is the cdf of $X_j$ for $i = 1,\ldots,p$, relies on proving asymptotic normality for the variables
$$\fY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\Z_t - \z)\fv(\Z_t, \frho_0),$$
where $\fv(\cdot) = \partial L(\cdot)/\partial\frho$ in the same way as in Section \@ref(app-loccor) in this supplement. This has been done by @otneim2017locally in the iid case using arguments from @schervish1995theory, and by @otneim2017conditional in the $\alpha$-mixing case, using arguments from @fan2008nonlinear. We need here to prove asymptotic normality of the variables

$$\hfY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\hZ_t - \z)\fv(\hZ_t,\frho_0),$$
with $\hnZ_t = \Phi^{-1}(F_n(\X_t))$, and $F_n(\X_t)$ is the application of the marginal empirical distribution functions to the corresponding components of the vector $\X_t$.

In order to ease notation we state the proof in the simplified case only, where the local correlations are estimated using *pairs* $(i,j)$ of variables. Then $Y_n$ can be taken to be a scalar. It can be extended to the trivariate case by adding the appropriate terms in the derivations below, without changing the final conclusion. Write $\hY_n(\z) = Y_n(\z) - (Y_n(\z) - \hY_n(\z))$ and do a Taylor expansion of $Y_n(\z)$ around $F_n(\X_t)$. Since we use product kernels, we then need derivatives of (assuming without loss of generality that $K(x) = K(-x)$)

@@

In the original version of this proof we used an argument based on the law of the iterated logarithm of @rio1995functional, leading to a rate of $n^{-1/2}(2 \log \log n )^{1/2}$, but an anynomous referee provided us with a much simpler argument based on a U-statistic decomposition for a strongly mixing process. In the following we omit the index $i=1,2$:

\begin{align*}
\frac{1}{n}\sum_{i=1}^n \left(F_n(X_{i,t}) - F(X_{i,t})\right) &= 
\frac{1}{n^2}\sum_{i=1}^n\sum_{j\neq i}^n \left(I(X_{j, t} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{i=1}^n \left(1 - F(X_{i,t}\right) \\
&= \frac{1}{n} \sum_{i=1}^n \left(0.5 - F(X_{i,t})\right) + O_p\left(n^{-1}\right) \\
&= O_p\left(n^{-1/2}\right) + O_p\left(n^{-1}\right) \\
&= O_p\left(n^{-1/2}\right).
\end{align*}

<!-- We need a bound on the supremum of $|F_n(X_t) - F(X_t)|$ and $|F_n(x) - F(x)|$ as $n \rightarrow \infty$. According to @arcones1995law, the best law of the iterated logarithm has been obtained by @rio1995functional. One statement of this is as follows: -->

<!-- Let $f(x) \in L_p$, $2<p\leq\infty$, $\sum\alpha_k k^{2/(p-2)} < \infty$, then the sequence $f(X_t)$ satisfies the compact LIL (*law of the iterated logarithms*), i.e. -->
<!-- $$\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{t=1}^n(f(X_t) - E[f(X_t)]) \right\}_{n=1}^{\infty}$$ -->
<!-- is relatively compact and its limit set is $[-\sigma, \sigma]$ where -->
<!-- $$\sigma^2 = \textrm{Var}\Big(f(X_t)\Big) + 2\sum_{s=1}^{\infty}\textrm{Cov}\Big(f(X_t), f(X_s)\Big).$$ -->
<!-- For a fixed $x$, take $f(X_t) = I(X_t\leq x)$. Then E$(f(X_t)) = F(x)$ and $f(x)\in L_p$ for all $p$, and -->
<!-- $$\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{s=1}^n \left(I(X_s \leq x) - F(x) \right) \right\}$$ -->
<!-- has its limit set in $[-\sigma, \sigma]$. Here  -->
<!-- $$\textrm{Var}(I(X_t \leq x)) = F(x)(1-F(x)) \leq \frac{1}{4},$$ -->
<!-- and according to Davydov's Lemma (see e.g. @fan2008nonlinear p. 278), -->
<!-- $$\big| \textrm{Cov}(I(X_t \leq x), I(X_s \leq x))\big| \leq 8 \|I(X_t \leq x)\|_p \|I(X_s \leq x)\|_q \alpha(|s-t|)^{1-p^{-1}-q^{-1}}, \,\,\, p,q>1, \,\,\, \frac{1}{p}+\frac{1}{q} < 1,$$ -->
<!-- with $\| I(X_t \leq x)\|_p = \left(\textrm{E}|I(X_t \leq x)|^p\right)^{1/p}$. We may for example take $p=q=4$, leading to $\| I(X_t \leq x) \|_4 = F(x)^{1/4} \leq 1$. -->

<!-- Since we have assumed exponential mixing (much weaker conditions suffice), then Rio's condition $\sum \alpha_k k^{2/(p-2)} < \infty$ is satisfied, and the bounds of $-\sigma$ and $\sigma$ can be made independent of $x$. -->

<!-- Having  -->
<!-- $$F_n(x) = \frac{1}{n}\sum_{t=1}^n I(X_t \leq x)$$ -->
<!-- means that -->
<!-- $$n^{\frac{1}{2}}[2\log \log n]^{-\frac{1}{2}} (F_n(x) - F(x))$$ -->
<!-- has its limits in $[-\sigma, \sigma]$ or that the supremum of $|F_n(x) - F(x)|$ is almost surely bounded by $n^{-1/2}(2\log\log n)^{1/2}$. -->

<!-- This immediately implies that -->
<!-- $$\frac{1}{n}\sum_{t=1}^n\left(F_n(X_t) - F(X_t)\right)$$ -->
<!-- is almost surely (and consequentially in probability) bounded by $n^{-\frac{1}{2}}(2 \log \log n)^{\frac{1}{2}}$ which for all conceivable bandwidths  is faster than the convergence rates of the CLT for $Y_n(\z)$, which is $(nb^2)^{-\frac{1}{2}}$. -->

@@

We can use the bootstrap theory by @lacal2018estimating more or less directly, which in turn is based on Gon\c{c}alves and White [-@gonccalves2002bootstrap; -@gonccalves2004maximum], but we spell out some details  below. First, we need to specify the analogue of the conditions A1--A5 by @lacal2018estimating, where we adapt their notation to our situation by replacing the parameter $\theta$ with $\rho_{ij}$, and the time series $\{X_t, Y_t\}$ by our multivariate observations $\{\Z_t\}$. Since $\rho = \rho_{ij}$ is estimated using the two corresponding data vectors only, we can, for the time being and without loss of generality, assume that the observation series is bivariate: ${\Z_t} = \{Z_{i,t}, Z_{j,t}\}$. Note further that it suffices to consider the true pseudo-observations $\{\Z_t\}$ instead of the empirical pseudo observations $\{\hZ_t\}$ in the following because the difference between them is asymptotically negligible according to Theorem 3.2 in the main article.

To state the bootstrap results we need a few more concepts and three additional conditions. We block-bootstrap the entire vector series $\{\Z_t\}$. This automatically leads to the same block-bootstrap for any pair of components which we have denoted by $\{Z_{i,t}, Z_{j,t}\}$ above. @lacal2018estimating use both the block bootstrap as well as the stationary bootstrap, and the results below hold irrespective of this particular choice. We will therefore not distinguish between them here, but rather refer the reader to @lacal2018estimating for further details on this distinction. Bootstrapped quantities will be denoted with a star ($*$) in the following.

@@

\begin{cor}
Assume the conditions of Theorem \ref{thm:E1}. Let $\alpha_{n,b}(i,j|\Z_3) \stackrel{\textrm{def}}{=} \widehat \rho(Z_{1i}, Z_{2j}|\Z_3)$ be the $i,j$th element of the estimated partial correlation matrix of $(\Z_1, \Z_2)$ given $\Z_3$. This depends on scalar coordinates $(z_1, z_2)$ and a vector coordinate $\z_3$. Note that $\alpha_{n,b}(i,j|\Z_3)$ is a continuous function of a vector of local correlations according to any of the alternatives (i)', (ii)--(v) in Section \ref{chap:estimation}. Then $\alpha_{n,b}(i,j|\Z_3) - \alpha^*_{n,b}(i,j|\Z_3) \buildrel {\mathbb{P}_{\omega}^{*},\mathbb{P}} \over {\longrightarrow} 0$. 
\label{cor:E3}
\end{cor}

@@

It should be carefully noted that it is not entirely obvious what the order of the term $C_n(A_b)$ is as $\bb = (b_i,b_j) = (b,b) \to \bm{0}$ and $n \to \infty$. Looking at the expressions \@ref(eq:t3) and \@ref(eq:t4), it is the term \@ref(eq:t3) that causes potential trouble due to the presence of $b_i^{-1}$ and $b_j^{-1}$ in the two kernel function terms in the product kernel. These are cancelled out by the integration over $(z_i,z_j)$, however. This means that the true convergence rate for $T_{n,b}-T$ is $n^{-1/2}$. This may be somewhat surprising since the convergence rate of $\frho_{n,b}(\z_{ij})-\frho(\z_{ij})$ is $(nb_1b_2)^{-1/2}$, but this is a well-known phenomenon for functionals of type $T_{n,b}$. In Lacal and Tjøstheim [-@lacal2018estimating, Remark 3.1] the convergence rate is changed from $(nb_1^3b_2^3)^{-1/2}$ to $(nb_1^2b_2^2)^{-1/2}$. These results are consistent with similar results for test functionals for independence in @joe1989estimation.

@@

Once a bootstrap realization $\{\Z_{1,t}^*,\Z_{2,t}^*,\Z_{3,t}^*\}$ is created, following @su2008nonparametric, these realizations are now taken as the original observations in the local likelihood estimation procedure. For a fixed bandwidth $b$ the population values, following the local Gaussian approach, can be taken as the solution of the integral equation (2) in @tjostheim2013local; see also corresponding equations in e.g. Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional] and eq. (19) in the main article. Note that in the present paper we have largely operated with two alternatives, the general vector case with inference via the pairwise local likelihood, or the the purely scalar case with three coordinates for the local Gaussian correlation. In both cases we have normalized the processes so that they have standard normal marginals.
